R package qtdshift: A functional programming approach to shifted probability distributions

### 2016-08-31 docs/article.Rmd
article

Shifted distributions are used daily in many statistical fields.   
A shifted distribution is a distribution to which is added a constant, $Y = X + a$, where $X$ and $Y$ are random variables and $a$ is a constant value.

The aim of this paper is to describe an R package, `qdtshift` which generates shifted distribution models for any distribution chosen. The packege was realized using a functional programming approach and it is ispired to Andrea Spano article "R package qdist: A functional programming approach to truncated probability ditributions", which provides an alternative method for computing from truncated probability distribution using a functional programming approach.

@@

* probability distribution functions: `p<dist>(q)` with `q` vector of quantiles;
* density functions: `d<dist>(x)` with `x` vector of quantiles;
* quantile functions: `q<dist>(p)` with `p` vector of probabilities;
* random number generation: `r<dist>(n)` with `n` number of observations.

@@

At first view, this approach seems easy and cool but thinking about it I realized that I would need to write a different function for each probability distribution I would like to work with. So I decided to use functional programming approach to develop four general functions which compute shifted distribution models for any distribution chosen:

@@

These functions are included in `qtdshift` package available at [https://github.com/veronicagiro/qshift](https://github.com/veronicagiro/qshift).

@@

They share the same logic: they take as input a probability distribution name, as a character string, and return the equivalent shifted distribution as a function object.

@@

We see in detail the computational process.   
Let us consider `dshift()`. As we seen, it takes as input the distribution name (`norm`) and proceeds as follows:

@@

Package `qtdshift` is based on two key concepts being part of the functional nature of the R programming language:

Theoretical references in chapters [Environments](http://www.quantide.com/website/ramarro-chapter-03/) and [Functions](http://www.quantide.com/website/ramarro-chapter-04/) of Andrea Spano weeb Book [Ramarro](http://www.quantide.com/website/ramarro-r-for-developers/), available on Quantide web site.

@@

    ```{r squnif_example}
    x <- c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)
not_shifted <- squnif(p = x, min = 6, max = 12)
positive_shifted <- squnif(p = x, min = 6, max = 12, shift = 1)
negative_shifted <- squnif(p = x, min = 6, max = 12, shift = -1)
    ```
    We can visualize these results by plotting them:

@@

In his article, Andrea Spanò use the newly defined probability distribution functions as a base for defining a function for maximum likelihood estimate for the truncated normal distribution. I demonstrate that it is possible also for shifted normal distribution.
Let us see how.

We start from the previously defined fucntion `sdnorm()`, that will be used within an estimator function:

```{r max_likelihood_est}
ltnorm <- function(x, shift = 0) {
  theta <- c(mean(x), sd(x))
  ll <- function(theta, x, shift = 0) {
    mean <- theta[1]
    sd <- theta[2]
    ld <- sdnorm(x = x, mean = mean, sd = sd, shift=shift, log=TRUE)
    -sum(ld)
    } 
  optim(par = theta, fn = ll, x = x, shift=shift)[["par"]]
}
```

```{r max_likelihood_est_example}
# generate a random sample from a shifted normal distribution
x <- srnorm(n = 1000, mean = 5, sd = 2, shift = 1)
# maximum likelihood estimate for the shifted normal distribution
ltnorm(x = x, shift  = 1)
```

@@

Spano A. .  _R package qdist: A functional programming approach to truncated probability ditributions_. url

>>>>

Shifted distributions are used daily in many statistical fields.   
A shifted distribution is a distribution with a constant added, $Y = X + a$, where $X$ and $Y$ are random variables and $a$ is a constant value.  
The aim of this paper is to describe an R package, `qdtshift` which generates shifted distribution models for any distribution chosen. The package was realized using a functional programming approach and inspires its computational method to Andrea Spanò article "R package qdist: A functional programming approach to truncated probability ditributions", which provides an alternative method for computing from truncated probability distribution using a functional programming approach.

@@

* density functions: `d<dist>(x)` with `x` vector of quantiles;
* probability distribution functions: `p<dist>(q)` with `q` vector of quantiles;
* quantile functions: `q<dist>(p)` with `p` vector of probabilities;
* random number generation: `r<dist>(n)` with `n` number of observations.

@@

At first view, this approach seems easy and cool but thinking about it, we realize that we would need to write a different function for each probability distribution we would like to work with. Functional programming approach could be the solution, as Andrea Spanò demonstrates in his article (see "R package qdist: A functional programming approach to truncated probability ditributions" for more details).

`qtdshift` package includes four general functions which compute shifted distribution models for any distribution chosen:

@@

`qtdshift` package is available at [https://github.com/veronicagiro/qshift](https://github.com/veronicagiro/qshift).

@@

The four functions share the same logic: they take as input a probability distribution name, as a character string, and return the equivalent shifted distribution as a function object.

@@

Talking about the computational process, we can say that package `qtdshift` is based on two key concepts being part of the functional nature of the R programming language:

For example, let us consider `dshift()`. As we seen, it takes as input the distribution name (`norm`) and proceeds as follows:

@@

Other references are in chapters [Environments](http://www.quantide.com/website/ramarro-chapter-03/) and [Functions](http://www.quantide.com/website/ramarro-chapter-04/) of Andrea Spanò weeb Book [Ramarro](http://www.quantide.com/website/ramarro-r-for-developers/), available on Quantide web site.

@@

    ```{r squnif_example}
x <- (1:999)/1000
not_shifted <- squnif(p = x, min = 6, max = 12)
positive_shifted <- squnif(p = x, min = 6, max = 12, shift = 1)
negative_shifted <- squnif(p = x, min = 6, max = 12, shift = -1)
    ```
    We can visualize these results by plotting them:

@@

Further implementations are possible starting from the definition of a shifted distribution.  
Suppose we want to define a function for maximum likelihood estimate for the shifted weibull distribution.

Firstly let us define the shifted density function for the weibull distribution:

```{r sdweibull_def}
sdweibull <- dshift("weibull")
```

and, subsequently, by using `sdweibull()` within an estimator function:

```{r max_likelihood_est}
ltweibull <- function(x){
  # starting points for parameters (scale, shape) and shift
  shift <- min(x) - 0.01
  x1 <- x - shift
  shape <- (sd(x1)/mean(x1))^(-1.086)
  scale <- mean(x1)/gamma(1+1/shape)
  # parameters vector definition
  theta <- c(shape, scale, shift)
  # likelihood function
  ll <- function(theta, x){
    shape <- theta[1]
    scale <- theta[2]
    shift <- theta[3]
    ld <- sdweibull(x = x, shape = shape, scale = scale, shift = shift, log = TRUE)
    -sum(ld)
  } 
  # maximum likelihood estimation
  optim(par = theta, fn = ll, x = x)[["par"]]
}
```

Let us see how it works on a random generated sample from a shifted weibull distribution:

```{r max_likelihood_est_example}
# generate a random sample from a shifted weibull distribution
srweibull <- rshift("weibull")
x <- srweibull(n = 100000, shape = 1, scale = 5, shift = 1)

# maximum likelihood estimate for the shifted weibull distribution
ltweibull(x = x)
```

@@

Spanò A. .  _R package qdist: A functional programming approach to truncated probability ditributions_. url
